Global code motion of congruent instructions

### 2017-11-06 cgo/paper.tex
undef

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP) which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail us. A part of
our implementation (aggressive code hoisting) is already merged
in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item a new optimistic algorithm to move congruent instructions out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

@@

\citet{click1995global} describes aggressive \gcm{} to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report a speedup of as high as $23\%$.

@@

\section{A fastt Code motion of congruent computations}
The algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers and the post-dominance frontiers of the CFG. These points, being
structural properties of a graph, do not change during code-motion.

@@

\subsubsection{Barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. In the absence of context, as in our current
implementation, some instructions which might be safe are still classified as
barriers, e.g., volatile loads/stores, calls with missing attributes.

>>>>

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP) which results in prohibitively
high compile time which is not suitable for industrial compilers. To the best of
our knowledge we have not found any reference which explored \gcm{} of congruent
instructions. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item a cost model to reduce interference and hence, reduce spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

@@

\citet{click1995global} describes aggressive \gcm{} in sparse representation. It
first schedules all the instructions as early as possible. This results in very
long live ranges which is mitigated by again scheduling all the instructions as
late as possible. They report a speedup of as high as $23\%$ but there is no
cost model to limit amount of scheduling and could often introduce redundant
computations. It is also not mentioned how safety is ensured while hoisting an
instruction when it is not anticipable. The legality checks are only based on
barriers (Section~\ref{subsec:barriers}) and availability of used operands. Even
introducing a simple integer addition can also result in undefined behavior
\cite{undef}.  Except from this paper we found no reference of global scheduling
in sparse representation.

@@

@@

\section{A fast algorithm for code motion congruent computations}
The algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers and the post-dominance frontiers of the CFG. These points, being
structural properties of a graph, do not change during code-motion.

@@

\subsubsection{Barriers}
\label{subsec:barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. A basic-block in \LLVM{} IR is actually an
extended basic-block because there might be non-returning calls in the middle of
a basic-block. Essentially, any instruction that cannot guarantee progress is
marked as a barrier. In the absence of context, as in our current
implementation, some instructions which might be safe are still classified as
barriers, e.g., volatile loads/stores, calls with missing attributes.
